Circular Regression Trees and Forests with an Application to Probabilistic Wind Direction Forecasting

### 2019-11-29 pkg/circtree/inst/wind_paper/draft/draft.Rnw
revision of introduction
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2903 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 2d0f5072b8edbdfe0e03245dccf1b82b60388198

additive effects and their interpretation are not intuitive for circular responses.
Due to the highly non-linear link function, the representation of smooth transitions on the unit circle
is not straightforward; additionally, the same rotation
can be obtained in either positive or negative direction on the circle.
%%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
%%%are built on additive regression models.
%%%Generally these provide a powerful toolbox for modeling particularly smooth effects.
%%%however, in the case of modeling circular data 
%%%additive effects conditional on linear covariates are not intuitive
%%%due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
%%%can be obtained in either positive or negative direction on the circle.
%a possibly highly non-linear link function
%can induce a more complex behavior of the modeled parameters as the location
%parameter shifts closely towards the ``origin point'' $0$, also depending on the 
%direction of the shift.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests 
do not require a pre-specified link function or any other model assumptions as covariates and their possible interactions are
selected automatically. Non-additive effects are captured by the tree structure while 
forests allow for modeling smooth changes as well. Thus, a high level of 
flexibility is provided while a full distributional model is specified in each terminal 
node offering a wide range of inference. Moreover, as covariates are employed for splitting only and not
considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.
Moreover, abrupt changes 
might not be detected properly owing to the enforced additive structure and selected model 
specification such as the form of linear combinations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression 
trees where binary splits are made based on an angular distance measure employed to evaluate node homogeneity.
%which is evaluated between 
%the observed response values and the conditional mean direction of the circular response variable
%and in that way serves as a measure of homogeneity within subgroups. 
%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%%subset of observations.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not only for 
linear but also for a circular covariate, the tree only considers effects in the mean direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.

>>>>

additive effects and their interpretation are not intuitive for circular responses
due to the highly non-linear link function. Consequently, the representation of smooth transitions 
on the unit circle is not straightforward. In addition, the same rotation can be obtained in either 
positive or negative direction on the circle leading to an ambiguous interpretation.

%%%All listed methods and most of their further developments \citep[see, e.g.,][]{Jammalamadaka+Sengupta:2001} 
%%%are built on additive regression models.
%%%Generally these provide a powerful toolbox for modeling particularly smooth effects.
%%%however, in the case of modeling circular data 
%%%additive effects conditional on linear covariates are not intuitive
%%%due to a possibly strongly non-linear rotation on the unit circle, whereby the same transformation
%%%can be obtained in either positive or negative direction on the circle.
%a possibly highly non-linear link function
%can induce a more complex behavior of the modeled parameters as the location
%parameter shifts closely towards the ``origin point'' $0$, also depending on the 
%direction of the shift.

As a very intuitive and data-driven alternative, we propose a flexible
tree-based regression approach for modeling circular data by applying the von
Mises distribution within the methodology of distributional trees and
distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
%In order to avoid these difficulties in regression modeling of circular data 
%(and also the requirement of initial values) 
%we propose an alternative and very flexible 
%approach by applying the von Mises distribution within the methodology of distributional 
%trees and distributional forests \citep{Schlosser+Hothorn+Stauffer:2019}. 
The resulting circular regression trees and forests avoid the discussed difficulties 
of circular GLMs by separating the data into more homogeneous subgroups and fitting a 
simple intercept-only model for each resulting subset.
Therefore, no covariates and, thus, no link function must be employed in the estimation
of the group-specific distribution parameters.
Moreover, by incorporating the framework of distributional modeling we allow for detecting 
splits in all distribution parameters and provide a fully specified circular response distribution 
for each terminal node offering a wide range of inference.
In addition, the employed tree structure captures non-additive effects while forests allow for 
modeling smooth changes as well.
Furhtermore, covariates and their possible interactions do not need to specified in advance as they are
selected automatically.

%Moreover, non-additive effects are captured by the tree structure while forests allow for 
%modeling smooth changes as well. Thus, a high level of flexibility is provided while a full 
%distributional model is specified in each terminal node offering a wide range of inference.

%The resulting circular regression trees and forests do not require a pre-specified link function 
%or any other model assumptions as covariates and their possible interactions are
%selected automatically.

%Moreover, as covariates are employed for splitting only and not
%considered for the parameter estimation within the resulting subgroups, their effects are easier to be interpreted for circular responses.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This novel approach to circular regression trees and forests complements the literature 
on tree-based circular modeling.
%In particular, 
\cite{Lund:2002} has already introduced the concept of circular regression trees where 
binary splits are made based on an angular distance measure employed to evaluate node 
homogeneity.
While this approach and also further extensions \citep{Larraondo+Inza+Lozano:2018} allow not 
only for linear but also for a circular covariate, the tree only considers effects in the mean 
direction for splitting, i.e., only the location parameter $\mu$ is regressed on covariates.
Therefore, the introduced distributional approach to circular regression trees extends the concept
of previous circular trees by considering splits in all distribution parameters and 
yielding a full probabilistic model. In addition, as a natural further development ensembles or forests 
of circular regression trees are presented in this study. 
% that can improve forecasts by regularizing and stabilizing the model.

%which is evaluated between 
%the observed response values and the conditional mean direction of the circular response variable
%and in that way serves as a measure of homogeneity within subgroups. 
%%For a circular response $\theta$, covariate(s) $z$ and the conditional mean direction $\mu(z)$
%%the employed circular distance is defined by $1-\cos(\theta - \mu(z)$.
%%In each terminal node a predicted values is provided by the sample mean direction of the corresponding 
%%subset of observations.
%Furthermore, even though an estimate of dispersion in the form of total node heterogeneity is provided 
%along with the predicted mean value in each terminal node
%By incorporating the framework of distributional modeling, we extend the idea of circular trees allowing splits in all distribution parameters and providing a fully specified circular response distribution for each terminal node.
